An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-29 comment.tex
removed efficiency

\textbf{Discussion}: In most pooling schemes, identifying an infected
individual requires positive results in multiple tests and hence
substantially increases false-negative rates. Furthermore, this
phenomenon is more pronounced when infection prevalence is low ---
exactly when pooling is most efficient. Thus, pooling presents an
inherent trade-off: it is most efficient when it is least
accurate. The deterioration of false-negative rates and the
aforementioned trade-off are inherent problems of pooling schemes and
should be kept in mind by practitioners and policy makers.
\end{abstract}

\section{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development of techniques to increase the the
throughput of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30,
  BayesianDorfman, MatrixPooling, LionDorfman}. These techniques,
often termed \emph{pooling}, or group testing, have originated in the
seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroguhly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

@@

\subsection{RT-PCR}\label{subsec:rtpcr}
We briefly explain the process of testing for SARS-CoV-2 by RT-PCR in
Section~\ref{subsec:rtpcr}. Our exposition intentionally avoids many
details, and a comprehensive review of RT-PCR can be found in
\cite{COVID-RTPCR, PCRBook}.

PCR is a process in which a targeted DNA strand's frequency is
amplified to create billions of copies in a reaction mixture. At the
end of this \emph{amplification} process, the presence of targeted DNA
molecules in a reaction mixture can be confidently
determined. Arguably, the best way to understand PCR is as a chain
reaction: Each targeted DNA molecule is copied into two, that are
copied into four and so forth. This cascade gives PCR its name:
polymerase chain reaction.

Specifically, the PCR process is comprised of cycles, with each cycle
doubling the abundance of the targeted DNA. In each cycle, each
double-strand DNA molecule is broken into two separate strands. An
enzyme called DNA polymerase generates a new double-strand DNA replica
from each separate strand. Replication cannot start without a specific
short DNA sequence, called a primer, that has to be introduced into
the reaction mixture. Introducing the right primer into the reaction
mixture ensures (almost) only the targeted DNA sequence is copied. A
protein is added to the reaction mixture, which emits light upon
successfully binding the targeted DNA. Once enough light is emitted
the tested reaction mixture is declared positive and a
\emph{detection} is said to occur. If, on the other hand, a predefined
number of cycles pass without a detection event then the reaction
mixture is declared negative.

Since the genetic material of SARS-CoV-2 is RNA rather than DNA, an
extra preprocessing step is required. An enzyme called reverse
transcriptase (hence the RT in RT-PCR) replicates existing RNA in the
reaction mixture into DNA. Once DNA replicas are made, the PCR process
proceeds as described above, targeting SARS-CoV-2 DNA replicas.

\subsection{Pooling}
One possible source of failed detection of SARS-CoV-2 RNA in pooled
RT-PCR testing (a \emph{false-negative}) is sample dilution. When
pooling, several samples are mixed, so the concentration of viral RNA
is reduced. This effect may cause a delay in amplification
\cite{DorfmanYuvalDor}, no detection and, consequently, a
false-negative. Another potential source for false-negatives is a
failed sampling process.

@@

\section{Methods}
We define and present the \emph{common assumption} in Section~\ref{subsec:common}
and refute it in Section~\ref{subsec:refute}. Our model is presented
in Section~\ref{subsec:ours}.

@@

\subsection{The common assumption}\label{subsec:common}
Previous studies of group testing strategies assumed that the
false-negative probability does not depend on the number of infected
samples, but merely on the existence of at least one such sample in a
pool \cite{Kim, OptimalDorfmanPool}. Current studies of pooling in the
context of SARS-CoV-2 also employ similar assumptions
\cite{Simplistic1, Simplistic2}. Specifically, these studies assume
that the probability of a negative result is the same for a pool with
a single sample coming from an infected individual and several
 samples from infected individuals. Explicitly:

@@

Below, we refer to \eqref{eq:common} as the \emph{common
  assumption}. The essence of the common assumption is that
amplification takes place for all samples at once.
%% We refute the assumptions underlying \eqref{eq:common} in
%% Section~\ref{subsec:ours} and suggest more plausible ones in
%% Section~\ref{subsec:ours}.

\subsection{Refuting the common assumption}\label{subsec:refute}
We refute the common assumption with experimental data summarized in
Table \ref{table}, collected from \cite{Salazar}. There, the authors
investigate Dorfman pooling and, regardless of the pooled test result,
follow up and test each pool member separately. We focus on 128 pools
for which at least one subsequent separate test was positive --- of
which 29 pooled tests were negative and 99 positive. In the data cited
in \cite{Salazar}, of the 29 negative pools, subsequent separate
testing yielded a single positive result in 24. In contrast, of the 99
positive pools, 42 yielded a single positive test upon subsequent
separate testing.

@@

\subsection{Our model}\label{subsec:ours}
Since the essence of the failed common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA occurs (or fails) for each sample
independently.  According to \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}, erroneous detection (a false-positive) does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our model with a small modification: We assume that
an erroneous amplification can occur in \emph{any} pool. Specifically,
it is possible that correct amplifications fails and an erroneous one
occurs simultaneously. This assumption is somewhat specific for the
current application of screening for SARS-CoV-2 via RT-PCR. For
example, cross-reactivity with other coronaviruses would have violated
this assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}.

@@

To summarize, we assume that for a single pool, a positive test result
is generated in one of two paths. Either, SARS-CoV-2 RNA from an
infected individual's sample is correctly amplified and detected, and
this can happen for each positive sample in a pool, with probability
$\Se$. Or, some erroneous amplification occurs (e.g. contaminant viral
RNA is introduced), an event that occurs at most once per pool, with
probability $1-\Sp$. Our model is illustrated in
Figure~\ref{fig:likelihood} and summarized in
%% A negative pooled test result occurs when all detection paths (both
%% correct and erroneous) fail. The probability of no false-detection
%% accounts for the $\spec$ term. The probability of no correct detection
%% is $(1-\sens)$ per infected individual. The probability that all such
%% paths fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}.%%  The probability of a positive result,
%% presented in \eqref{eq:likelihood b}, is simply the complement.
%% The probability of a negative pooled test result is presented in
%% \eqref{eq:likelihood a} (along with its complement
%% \eqref{eq:likelihood b}), and explained below.
%\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum_i\theta_i}. \label{eq:likelihood a}
    %
    %
    %
    %% \Pr(\data=1 | \theta) &= 1 - \spec\prod_{i=1}^N
    %% (1-\sens)^{\theta_i} = 1 - \spec(1-\Se)^{\sum\theta_i} . \label{eq:likelihood b}
  %\end{split}
\end{align}
%\end{subequations}
%% Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
%% recalling that $\data \in \{0,1\}$ yields:

\subsection{Scheme efficiency}
The efficiency of Dorfman pooling scheme is the expected number of
tests utilized. Let $T$ the number of tests. Then:
\begin{align}
  \begin{split}
    \mathbb{E}[T] &= 1 + N\Pr(\data=1)\\
    %
    %
    %
    &= 1 + N\sum_{\theta} \Pr(\theta) \Pr(\data=1|\theta) \\
    %
    %
    %
    &= 1 + N\sum_{k=0}^N \Pr(\sum \theta_i = k) \Pr(\data=1|\sum \theta_i = k) \\
    %
    %
    %
    &= 1 + N\sum_{k=0}^N \binom{N}{k}q^k(1-q)^k (1-\Sp(1-\Se)^k) \\
    %
    %
    %
    &= 1+N - N\Sp\sum_{k=0}^N \binom{N}{k}q^k(1-q)^k (1-\Se)^k \\
    %
    %
    %
    %&= 1+N - N\Sp \sum_{k=0}^N \binom{N}{k}(q-q\Se)^k(1-q)^k \\
    %
    %
    %
    &= 1+N - N\Sp (1-q\Se)^N\\    
  \end{split}
\end{align}

Compare to the common assumption:

\begin{align}
  \begin{split}
    \mathbb{E}[T] &= 1 + N\Pr(\data=1)\\
    %
    %
    %
    &= 1 + N\sum_{\theta} \Pr(\theta) \Pr(\data=1|\theta) \\
    %
    %
    %
    &= 1 + N\Big[\Pr(\theta_i = 0,\ \forall i)(1-\Sp) + \Pr(\exists i\ \theta_i =1)\Se\Big] \\
    %
    %
    %
    &= 1 + N\Big [(1-q)^N(1-\Sp) + (1-(1-q)^N)\Se \Big ] \\
    %
    %
    %
    &= 1 + N\Big [(1-q)^N(1-\Sp - \Se) + \Se \Big ] \\
    %
    %
    %
    &= 1 + \Se N + N(1-q)^N(1-\Sp - \Se).
  \end{split}
\end{align}

The difference is:
\begin{equation}
  \mathbb{E}^{\text{exact}}[T] - \mathbb{E}^{\text{common}}[T] = N \Big [1 - \Sp(1-q\Se)^N - \Se - (1-q)^N(1-\Sp - \Se) \Big ]
\end{equation}

\subsection{Scheme false-negative rate}
We calculate the false-negative rate for a single \emph{infected}
individual, henceforth referred to as "Donald" under our model and
under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Rephrasing, we wish to answer the following
question: what is the probability of not identifying Donald as
infected under a Dorfman pooling scheme?

@@

\begin{align}
  \begin{split}
    \Pr(\data=0|\theta_1=1) &= \sum_{\theta_2,\dots,\theta_N}
    \Pr(\theta_2,\dots,\theta_N)
    \Pr(\data=0|\theta_1=1,\theta_2,\dots,\theta_N) \\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\Pr(\sum_{i=2}^N\theta_i=k)
    \Pr(\data=0|\theta_1=1,\sum_{i=2}^N\theta_i = k)\\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\binom{N-1}{k}q^k(1-q)^{N-1-k} \Sp(1-\Se)^{1+k}\\
    %
    %
    %
    &= \Sp(1-\Se) \sum_{k=0}^{N-1}\binom{N-1}{k}
    \left(q(1-\Se)\right)^k(1-q)^{N-1-k}\\
    %
    %
    %
    &= \Sp(1-\Se)(1-q\Se)^{N-1}, \text{ by Newton's Binomial Theorem.} \\
  \end{split}
\end{align}

@@

\subsubsection{Comparison metric}
The single test false-negative rate $1-\Se$ and scheme false-negative
rate $\mi$ are compared via:
\begin{equation}\label{eq:erel}
\err{exact} := \frac{\mi - (1-\Se)}{1-\Se} \cdot 100.
\end{equation}
$\err{exact}$ is the percentage increase in the pooling scheme false-negative
rate, relative to the single test false-negative rate.

@@

\section{Results}\label{section:results}
\subsection{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\err{exact}$ can be expected (Figure \ref{fig1}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure \ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\err{common}$ and $\err{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{fig1}
\end{figure}

\subsection{Scheme efficiency}
Our results suggest that the efficiency under the common assumption is
always greater than the efficiency under our exact model. This should
not come as a surprise, as the common assumption implies larger
false-negative rates and, consequently, a lower number of tests being
utilized. As can be expected, the discrepancy is largest for lower
test sensitivity rates $\Se$.

\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_efficiency.jpg}
  \caption{Relative increase in Dorfman pooling efficiency
    $\eff{exact}$. Left: Colors represent $\eff{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\eff{common}$ and $\eff{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{fig2}
\end{figure}

\section{Discussion}

\subsection{Short recap}
We have presented a novel model for SARS-CoV-2 pooled tests. To the best of our knowledge, ours is the
only one that is actually based on experimental data.

@@

\subsection{Implications for group testing in different prevalence settings}
Our results have practical implications for the use of pooling
schemes. As the COVID-19 pandemic progresses, the infection prevalence
in various tested populations undergoes frequent changes. Hence, as
our results suggest, pooling strategies for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers
\cite{JTM,RobinHood}, potentially importing infections, or
asymptomatic individuals, such as children, and can be crucial for
controlling outbreaks \cite{MinaScience}.

@@

\subsection{Limitations}
Lacking large amounts of experimental data, it is hard to assess the
validity of our model. The data in \cite{Salazar} does allow us to
reject the common assumption but not much more. Our model is as simple
as possible, given that the standard assumptions are refuted and
dilution effects are not taken into consideration.

\subsection{Ways to reduce errors}
Although our application focused on Dorfman pooling, our analysis
applies to other pooling schemes. Pooling schemes
(e.g. \cite{MatrixPooling,Lion, Kim}), require some sequence of
positive pooled results to correctly identify Donald as
infected. Consider the hierarchical pooling scheme \cite{Lion, Kim}:
If the first pool yields a positive result, it is split in two. Then
the splitting is repeated until resulting pools are negative or
individuals are tested separately. With an initial pool size of 32,
Donald will necessarily have to test positive in pools of size 32, 16,
8, 4 and 2, as well as in a single test, for the scheme to correctly
identify him as infected. Compare this to the Dorfman scheme that
requires a positive test in a pool of size $N=8$, and an additional
single positive test to identify Donald as infected. The hierarchical
pooling scheme of \cite{Lion, Kim} will necessarily yield more
false-negatives than Dorfman pooling --- there are additional places
for it to fail.

@@

\subsection{Conclusion}
We have no doubt that further study of the assumptions underlying
pooling is due. Furthermore, adapting common group testing algorithms
\cite{Kim} to novel error models should be further studied.

@@

%\bibliographystyle{amsplain}
\bibliographystyle{ama}
\bibliography{refs}

>>>>

\textbf{Discussion}: In most pooling schemes, identifying an infected
individual requires positive results in multiple tests and hence
substantially increases false-negative rates. Furthermore, this
phenomenon is more pronounced when infection prevalence is low ---
exactly when pooling is most efficient. Thus, pooling presents an
inherent trade-off: it is most efficient when it is least
accurate. The deterioration of false-negative rates and the
aforementioned trade-off are inherent problems of pooling schemes and
should be kept in mind by practitioners and policy makers.
\end{abstract}
\newpage

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development of techniques to increase the the
throughput of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30,
  BayesianDorfman, MatrixPooling, LionDorfman}. These techniques,
often termed \emph{pooling}, or group testing, have originated in the
seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroguhly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

@@

% \subsection{RT-PCR}\label{subsec:rtpcr}
% We briefly explain the process of testing for SARS-CoV-2 by RT-PCR in
% Section~\ref{subsec:rtpcr}. Our exposition intentionally avoids many
% details, and a comprehensive review of RT-PCR can be found in
% \cite{COVID-RTPCR, PCRBook}.

% PCR is a process in which a targeted DNA strand's frequency is
% amplified to create billions of copies in a reaction mixture. At the
% end of this \emph{amplification} process, the presence of targeted DNA
% molecules in a reaction mixture can be confidently
% determined. Arguably, the best way to understand PCR is as a chain
% reaction: Each targeted DNA molecule is copied into two, that are
% copied into four and so forth. This cascade gives PCR its name:
% polymerase chain reaction.

% Specifically, the PCR process is comprised of cycles, with each cycle
% doubling the abundance of the targeted DNA. In each cycle, each
% double-strand DNA molecule is broken into two separate strands. An
% enzyme called DNA polymerase generates a new double-strand DNA replica
% from each separate strand. Replication cannot start without a specific
% short DNA sequence, called a primer, that has to be introduced into
% the reaction mixture. Introducing the right primer into the reaction
% mixture ensures (almost) only the targeted DNA sequence is copied. A
% protein is added to the reaction mixture, which emits light upon
% successfully binding the targeted DNA. Once enough light is emitted
% the tested reaction mixture is declared positive and a
% \emph{detection} is said to occur. If, on the other hand, a predefined
% number of cycles pass without a detection event then the reaction
% mixture is declared negative.

% Since the genetic material of SARS-CoV-2 is RNA rather than DNA, an
% extra preprocessing step is required. An enzyme called reverse
% transcriptase (hence the RT in RT-PCR) replicates existing RNA in the
% reaction mixture into DNA. Once DNA replicas are made, the PCR process
% proceeds as described above, targeting SARS-CoV-2 DNA replicas.

%\subsection{Pooling}
% One possible source of failed detection of SARS-CoV-2 RNA in pooled
% RT-PCR testing (a \emph{false-negative}) is sample dilution. When
% pooling, several samples are mixed, so the concentration of viral RNA
% is reduced. This effect may cause a delay in amplification
% \cite{DorfmanYuvalDor}, no detection and, consequently, a
% false-negative. Another potential source for false-negatives is a
% failed sampling process.

@@

\section*{Methods}
% We define and present the \emph{common assumption} in Section~\ref{subsec:common}
% and refute it in Section~\ref{subsec:refute}. Our model is presented
% in Section~\ref{subsec:ours}.

@@

\subsection*{The common assumption}\label{subsec:common}
Previous studies of group testing strategies assumed that the
false-negative probability does not depend on the number of infected
samples, but merely on the existence of at least one such sample in a
pool \cite{Kim, OptimalDorfmanPool}. Current studies of pooling in the
context of SARS-CoV-2 also employ similar assumptions
\cite{Simplistic1, Simplistic2}. Specifically, these studies assume
that the probability of a negative result is the same for a pool with
a single sample coming from an infected individual and several
 samples from infected individuals. Explicitly:

@@

Below, we refer to \eqref{eq:common} as the \emph{common
  assumption}. 
%% We refute the assumptions underlying \eqref{eq:common} in
%% Section~\ref{subsec:ours} and suggest more plausible ones in
%% Section~\ref{subsec:ours}.

\subsection*{Refuting the common assumption}\label{subsec:refute}
We refute the common assumption with experimental data summarized in
Table \ref{table}, collected from \cite{Salazar}. There, the authors
investigate Dorfman pooling and, regardless of the pooled test result,
follow up and test each pool member separately. We focus on 128 pools
for which at least one subsequent separate test was positive --- of
which 29 pooled tests were negative and 99 positive. In the data cited
in \cite{Salazar}, of the 29 negative pools, subsequent separate
testing yielded a single positive result in 24. In contrast, of the 99
positive pools, 42 yielded a single positive test upon subsequent
separate testing.

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA occurs (or fails) for each sample
independently. According to \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our model with a small modification: We assume that
an erroneous amplification can occur in \emph{any} pool. Specifically,
it is possible that correct amplifications fails and an erroneous one
occurs simultaneously. This assumption is somewhat specific for the
current application of screening for SARS-CoV-2 via RT-PCR. For
example, cross-reactivity with other coronaviruses would have violated
this assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}.
% To summarize, we assume that for a single pool, a positive test result
% is generated in one of two paths. Either, SARS-CoV-2 RNA from an
% infected individual's sample is correctly amplified and detected, and
% this can happen for each positive sample in a pool, with probability
% $\Se$. Or, some erroneous amplification occurs (e.g. contaminant viral
% RNA is introduced), an event that occurs at most once per pool, with
% probability $1-\Sp$.
These assumptions lead to our model, which is illustrated in 
Figure~\ref{fig:likelihood} and summarized in
%% A negative pooled test result occurs when all detection paths (both
%% correct and erroneous) fail. The probability of no false-detection
%% accounts for the $\spec$ term. The probability of no correct detection
%% is $(1-\sens)$ per infected individual. The probability that all such
%% paths fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}.%%  The probability of a positive result,
%% presented in \eqref{eq:likelihood b}, is simply the complement.
%% The probability of a negative pooled test result is presented in
%% \eqref{eq:likelihood a} (along with its complement
%% \eqref{eq:likelihood b}), and explained below.
%\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum_i\theta_i}. \label{eq:likelihood a}
    %
    %
    %
    %% \Pr(\data=1 | \theta) &= 1 - \spec\prod_{i=1}^N
    %% (1-\sens)^{\theta_i} = 1 - \spec(1-\Se)^{\sum\theta_i} . \label{eq:likelihood b}
  %\end{split}
\end{align}
%\end{subequations}
%% Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
%% recalling that $\data \in \{0,1\}$ yields:

@@

%% \subsection{Scheme efficiency}
%% The efficiency of Dorfman pooling scheme is the expected number of
%% tests utilized. Let $T$ the number of tests. Then:
%% \begin{align}
%%   \begin{split}
%%     \mathbb{E}[T] &= 1 + N\Pr(\data=1)\\
%%     %
%%     %
%%     %
%%     &= 1 + N\sum_{\theta} \Pr(\theta) \Pr(\data=1|\theta) \\
%%     %
%%     %
%%     %
%%     &= 1 + N\sum_{k=0}^N \Pr(\sum \theta_i = k) \Pr(\data=1|\sum \theta_i = k) \\
%%     %
%%     %
%%     %
%%     &= 1 + N\sum_{k=0}^N \binom{N}{k}q^k(1-q)^k (1-\Sp(1-\Se)^k) \\
%%     %
%%     %
%%     %
%%     &= 1+N - N\Sp\sum_{k=0}^N \binom{N}{k}q^k(1-q)^k (1-\Se)^k \\
%%     %
%%     %
%%     %
%%     %&= 1+N - N\Sp \sum_{k=0}^N \binom{N}{k}(q-q\Se)^k(1-q)^k \\
%%     %
%%     %
%%     %
%%     &= 1+N - N\Sp (1-q\Se)^N\\    
%%   \end{split}
%% \end{align}

%% Compare to the common assumption:

%% \begin{align}
%%   \begin{split}
%%     \mathbb{E}[T] &= 1 + N\Pr(\data=1)\\
%%     %
%%     %
%%     %
%%     &= 1 + N\sum_{\theta} \Pr(\theta) \Pr(\data=1|\theta) \\
%%     %
%%     %
%%     %
%%     &= 1 + N\Big[\Pr(\theta_i = 0,\ \forall i)(1-\Sp) + \Pr(\exists i\ \theta_i =1)\Se\Big] \\
%%     %
%%     %
%%     %
%%     &= 1 + N\Big [(1-q)^N(1-\Sp) + (1-(1-q)^N)\Se \Big ] \\
%%     %
%%     %
%%     %
%%     &= 1 + N\Big [(1-q)^N(1-\Sp - \Se) + \Se \Big ] \\
%%     %
%%     %
%%     %
%%     &= 1 + \Se N + N(1-q)^N(1-\Sp - \Se).
%%   \end{split}
%% \end{align}

%% The difference is:
%% \begin{equation}
%%   \mathbb{E}^{\text{exact}}[T] - \mathbb{E}^{\text{common}}[T] = N \Big [1 - \Sp(1-q\Se)^N - \Se - (1-q)^N(1-\Sp - \Se) \Big ]
%% \end{equation}

\subsection*{Scheme false-negative rate}
We calculate the false-negative rate for a single \emph{infected}
individual, henceforth referred to as "Donald" under our model and
under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Equivalently, what is the probability of not identifying Donald as
infected under a Dorfman pooling scheme?

@@

\begin{align}
  \begin{split}
    \Pr(\data=0|\theta_1=1) &= \sum_{\theta_2,\dots,\theta_N}
    \Pr(\theta_2,\dots,\theta_N)
    \Pr(\data=0|\theta_1=1,\theta_2,\dots,\theta_N) \\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\Pr(\sum_{i=2}^N\theta_i=k)
    \Pr(\data=0|\theta_1=1,\sum_{i=2}^N\theta_i = k)\\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\binom{N-1}{k}q^k(1-q)^{N-1-k} \Sp(1-\Se)^{1+k}\\
    %
    %
    %
    &= \Sp(1-\Se) \sum_{k=0}^{N-1}\binom{N-1}{k}
    \left(q(1-\Se)\right)^k(1-q)^{N-1-k}\\
    %
    %
    %
    &= \Sp(1-\Se)(1-q\Se)^{N-1}. \\
  \end{split}
\end{align}

@@

\subsubsection*{Comparison metric}
The single test false-negative rate $1-\Se$ and scheme false-negative
rate $\mi$ are compared via:
\begin{equation}\label{eq:erel}
\err{exact} := \frac{\mi - (1-\Se)}{1-\Se} \cdot 100.
\end{equation}
$\err{exact}$ is the percentage increase in the pooling scheme false-negative
rate, relative to the single test false-negative rate.

@@

\section*{Results}\label{section:results}
\subsection*{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\err{exact}$ can be expected (Figure \ref{fig1}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure \ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\err{common}$ and $\err{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{fig1}
\end{figure}

\section*{Discussion}
%\subsection{Short recap}
We have presented a novel model for SARS-CoV-2 pooled tests. To the best of our knowledge, ours is the
only one that is actually based on experimental data.

@@

%\subsection{Implications for group testing in different prevalence settings}
Our results have practical implications for the use of pooling
schemes. As the COVID-19 pandemic progresses, the infection prevalence
in various tested populations undergoes frequent changes. Hence, as
our results suggest, pooling strategies for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers
\cite{JTM,RobinHood}, potentially importing infections, or
asymptomatic individuals, such as children, and can be crucial for
controlling outbreaks \cite{MinaScience}.

@@

%\subsection{Limitations}
Lacking large amounts of experimental data, it is hard to assess the
validity of our model. The data in \cite{Salazar} does allow us to
reject the common assumption but not much more. Our model is as simple
as possible, given that the standard assumptions are refuted and
dilution effects are not taken into consideration.

%\subsection{Ways to reduce errors}
Although our application focused on Dorfman pooling, our analysis
applies to other pooling schemes. Pooling schemes
(e.g. \cite{MatrixPooling,Lion, Kim}), require some sequence of
positive pooled results to correctly identify Donald as
infected. Consider the hierarchical pooling scheme \cite{Lion, Kim}:
If the first pool yields a positive result, it is split in two. Then
the splitting is repeated until resulting pools are negative or
individuals are tested separately. With an initial pool size of 32,
Donald will necessarily have to test positive in pools of size 32, 16,
8, 4 and 2, as well as in a single test, for the scheme to correctly
identify him as infected. Compare this to the Dorfman scheme that
requires a positive test in a pool of size $N=8$, and an additional
single positive test to identify Donald as infected. The hierarchical
pooling scheme of \cite{Lion, Kim} will necessarily yield more
false-negatives than Dorfman pooling --- there are additional places
for it to fail.

@@

%\subsection{Conclusion}
We have no doubt that further study of the assumptions underlying
pooling is due. Furthermore, adapting common group testing algorithms
\cite{Kim} to novel error models should be further studied.

@@

\bibliographystyle{amsplain}
\bibliography{refs}
